Retrofitting Effect Handlers onto OCaml

### 2021-03-15 retro-concurrency.tex
Updates

Recall that OCaml programs are written with the expectation that function calls
return exactly once (\S\ref{sec:refine}). Consider the scenario when a
continuation is neither continued nor discontinued. Since fibers are
|malloc|ed, and are |free|d when the computation being handled returns a value
or raises an exception, not resuming continuations leaks memory. In addition,
unresumed continuations may also leak other resource such as open file
descriptors.

We consider well-behaved Multicore OCaml programs to resume their continuations
\emph{exactly once}. This can be enforced by installing a finaliser on every
captured continuation that calls |discontinue| on the continuation during
finalisation and ignores the result:
\begin{lstlisting}
Gc.finalise (fun k ->
  try ignore (discontinue k Unwind) with _ -> ()) k
\end{lstlisting}
However, Multicore OCaml does not enforce this behaviour in order to avoid the
costs associated with installing a finaliser every time a continuation is
captured.

For stack walking during GC, the only change necessary was to make the
runtime aware of fiber layout and how to follow the parent pointers. The
stack map generation and scanning each frame for roots remains the same. The
challenges and solutions for integrating fibers with the concurrent
mark-and-sweep GC of Multicore OCaml have been discussed
previously~\cite{Sivaramakrishnan20}.

@@

In this section, we evaluate the performance of Multicore OCaml effect handlers
against the performance requirements set in \S\ref{sec:req}. Multicore OCaml is
an extension of the 4.10.0 OCaml compiler with support for shared memory
parallelism and effect handlers. These results were obtained on a 2-socket
Intel\textregistered Xeon\textregistered Gold 5120
x86-64~\cite{IntelXeonGold5120Spec} server running Ubuntu 18.04 with 64GB of main
memory.

@@

Next we look at benchmarks that utilize non-local control flow using effect
handlers. First, we quantify the overhead of individual operations in effect
handling. Consider the following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, with 3 warm-up runs. We observed that the
sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
\textbf{23 ns, 5 ns, 11 ns and 7 ns}, respectively. For reference, the idle
memory load latency for the local NUMA node is 93.2 ns as measured using the
Intel MLC tool~\cite{mlc}.

@@

Generators allow data structures to be traversed on demand. Many languages
including JavaScript and Python provide generators as a built-in primitive.
Using effect handlers (|MC|), given any data structure (|'a t|) and its
iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can derive its
generator function (|val next : unit -> 'a option|). The code is available in
the supplementary material. We evaluate the performance of traversing a
complete binary tree of depth 25 using this generator. This involves $2^{26}$
stack switches in total. For comparison, we implemented a hand-written,
selective CPSed~\cite{Nielson01}, defunctionalised~\cite{Danvy01} version
(|cps|) and a concurrency monad (|monad|) version of the generator for the
tree. Both |cps| and |monad| versions are specialised to the binary tree with
the usual caveats of not using the stack for function calls. We observed that
the |cps| version was the fastest, thanks to specialisation and hand
optimisation. |MC| version was only 2.76$\times$ slower than |cps| while being
a generic solution, and the |monad| version was 8.69$\times$ slower than |cps|.

>>>>

Recall that OCaml programs are written with the expectation that function calls
return exactly once (\S\ref{sec:refine}). Consider the scenario when a
continuation is never resumed. Since fibers allocate memory for the stack using
|malloc|, which are |free|d when the computation returns, not resuming
continuations leaks memory. In addition, unresumed continuations may also leak
other system resources such as sockets and open file descriptors.

We make a pragmatic trade-off and expect the user code to resume captured
continuations \emph{exactly once}. One can use the GC support to free up
resources by installing a finaliser that discontinues the continuation and
ignores the result:

\begin{lstlisting}
Gc.finalise (fun k ->
  try ignore (discontinue k Unwind) with _ -> ()) k
\end{lstlisting}

This frees up both the memory allocated for the fiber stack as well as other
system resources, assuming that user code does not handle |Unwind| exception
and fails to re-raise it. But installing a finaliser on every captured
continuation by default introduces significant overhead to continuation
capture. Hence, we choose not to implement such a finaliser by default. It is
also useful to note that even if the memory for the fiber stack is managed by
the GC rather than being allocated in the C heap as we do now, we would still
need a finalisation mechanism to unwind the stack and release other system
resources that may be held by the continuation.

For stack walking during GC in order to find the roots, the only change
necessary was to make the runtime aware of the fiber layout and how to follow
the parent pointers. The stack map generation and scanning each frame for roots
remains the same. The challenges and the solutions for integrating fibers with
the concurrent mark-and-sweep GC of Multicore OCaml have been discussed
previously~\cite{Sivaramakrishnan20}.

@@

In this section, we evaluate the performance of Multicore OCaml effect handlers
against the performance requirements set in \S\ref{sec:req}. Multicore OCaml is
an extension of the 4.10.0 OCaml compiler with support for shared memory
parallelism and effect handlers. Since our objective is to evaluate the impact
of effect handlers, none of our benchmarks utilise parallelism. These results
were obtained on a 2-socket Intel\textregistered Xeon\textregistered Gold 5120
x86-64~\cite{IntelXeonGold5120Spec} server running Ubuntu 18.04 with 64GB of
main memory.

@@

Next we look at benchmarks that utilize non-local control flow using effect
handlers. First, we quantify the cost of individual operations in effect
handling. Consider the following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, with 3 warm-up runs. For calibration, the
idle memory load latency for the local NUMA node is 93.2 ns as measured using
the Intel MLC tool~\cite{mlc}. We observed that the sequences \texttt{a-b},
\texttt{b-c}, \texttt{c-d} and \texttt{d-e} took \textbf{23 ns, 5 ns, 11 ns and
7 ns}, respectively. The time in the sequence \texttt{a-b} is dominated by the
memory allocation. Thus, the individual operations in effect handling are fast.

@@

Generators allow data structures to be traversed on demand. Many languages
including JavaScript and Python provide generators as a built-in primitive.
Using effect handlers (|MC|), given any data structure (|'a t|) and its
iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can derive its
generator function (|val next : unit -> 'a option|)\footnote{https://gist.github.com/kayceesrk/eb0ab496c22861f21b1d9484772e982d}.
We evaluate the performance of traversing a complete binary tree of depth 25
using this generator. This involves $2^{26}$ stack switches in total. For
comparison, we implemented a hand-written, selective CPSed~\cite{Nielson01},
defunctionalised~\cite{Danvy01} version (|cps|) and a concurrency monad
(|monad|) version of the generator for the tree. Both |cps| and |monad|
versions are specialised to the binary tree with the usual caveats of not using
the stack for function calls. We observed that the |cps| version was the
fastest, thanks to specialisation and hand optimisation. |MC| version was only
2.76$\times$ slower than |cps| while being a generic solution, and the |monad|
version was 8.69$\times$ slower than |cps|.
